# Reproducible Research: Peer Assessment 1
by Chris Keen
========================================================

This is the R Markdown document I am submitting for Peer Assessment 1. 

## Loading and preprocessing the data
I used the unzip() function to unzip the file. Then created the table 'ActivityData' from the read.csv function.

```{r}
unzip("activity.zip")
ActivityData <- read.csv("activity.csv")

head(ActivityData)
str(ActivityData)
dim(ActivityData)
summary(ActivityData)

```

## What is mean total number of steps taken per day?

I used tapply to find the total number(sum) of steps taken each day.

```{r}
sumAD <- with(ActivityData, tapply(steps, date, sum, na.rm=TRUE))

```

Then I created a histogram using the results.
```{r fig.width=7, fig.height=6}
hist(sumAD, ylim=range(0:30), main="Daily Total Steps Taken", 
     xlab="Total of Steps")

```

Finally, I calculated the Mean and Median for the total steps taken each day.
```{r}
mnADsum <- mean(sumAD)
medADsum <- median(sumAD)

print(paste("ActivityData Mean: ", mnADsum))
print(paste("ActivityData Median: ", medADsum))

```

## What is the average daily activity pattern?
I calculated the Mean of steps for each interval across all the days.

```{r}
mnADintervals <- with(ActivityData, tapply(steps, interval, mean, na.rm=TRUE))

```
Next, I created a time series plot of the average number of steps taken at each 5-minute interval each day.

```{r}
plot(mnADintervals, type="l", xlab="5-Minute Intervals(24hr clock)", 
     ylab="Avg Number of Steps", main="Avg. Number of Steps per Interval", 
     xaxt="n")
    axis(side=1, at=seq(0,288, 12), labels=c(0:24))

```

Finally, I searched for the interval with the maximum number of steps on average for each day.

```{r}
maxADint <- max(mnADintervals)
names(mnADintervals)[which(mnADintervals == maxADint)]

```


## Imputing missing values
Calculate the number of 'NA' values in the dataset.
```{r}
sumADna <- sum(is.na(ActivityData))
print(sumADna)

```
I copied the ActivityData data frame to a new data frame, completeAD, in order to replace the 'NA' values. Then used a for loop to replace the 'NA' values with the means for each of the 5-minute intervals.  
```{r}
completeAD <- ActivityData

for (i in which(sapply(completeAD, is.numeric))) {
    completeAD[is.na(completeAD[, i]), i] <- 
        with(completeAD, tapply(steps, interval, mean, na.rm=TRUE))
}
```

I used tapply to find the total number(sum) of steps taken each day for the new dataset 'completeAD'.
```{r}
compADsteps <- with(completeAD, tapply(steps, date, sum, na.rm=TRUE))

```
Here is a comparison of the two datasets 'ActivityData', which has the NA values included and 'completeAD', which has the NA values replaced with the means for each of the 5-minute intervals.  
The first comparison is in the form of histograms
```{r}
par(mfcol=c(1,2))
hist(sumAD, ylim=range(0:40), 
     main="Daily Total Steps Taken\n (w/ NA's)", 
     xlab="Total of Steps")
hist(compADsteps, ylim=range(0:40), 
     main="Daily Total Steps Taken\n (NA's replaced)", 
     xlab="Total of Steps")

```

Finally, I created a data frame with the Mean and Median for each data set.
```{r}
mncompAD <- mean(compADsteps)
medcompAD <- median(compADsteps)

bothMN <- c(mnADsum, mncompAD)
bothMED <- c(medADsum, medcompAD)
difMNMED <- c((mncompAD-mnADsum), (medcompAD-medADsum))
MNMEDdf <- data.frame(bothMN, bothMED)
MNMEDdf <- rbind(MNMEDdf, difMNMED)
rownames(MNMEDdf) <- c("w/ NA's", "NA's replaced", "Difference")
colnames(MNMEDdf) <- c("Mean", "Median")
print(MNMEDdf)

```

## Are there differences in activity patterns between weekdays and weekends?
Created a new column in completeAD dataset for two level factor "weekday" and "weekend".

```{r}
completeAD$date <- as.Date(completeAD$date)
completeAD$weekdays <- as.factor(weekdays(completeAD$date))

levels(completeAD$weekdays) <- c("weekday", "weekday", "weekday", "weekday", 
                                "weekday", "weekend", "weekend")

str(completeAD)

```

Then I calculated the Mean of steps for each interval across the two factors.
Here I split the dataset 'completeAD' into two separate datasets; 'weekdayDF' and 'weekendDF'.  

```{r}
weekdayDF <- subset(completeAD, completeAD$weekdays == "weekday")
weekendDF <- subset(completeAD, completeAD$weekdays == "weekend")

```

This is the mean of steps for each interval for the "weekday" factor.
```{r}
mnWeekdayInts <- with(weekdayDF, tapply(steps, interval, mean, na.rm=TRUE))

```
This is the mean of steps for each interval for the "weekend" factor.
```{r}
mnWeekendInts <- with(weekendDF, tapply(steps, interval, mean, na.rm=TRUE))

```

Finally, I created two time plots to show comparison between the "Weekday" and "Weekend" Mean of steps for each interval.
```{r}
par(mfrow = c(2,1))
plot(mnWeekdayInts, type="l", xlab="5-Minute Intervals(24hr clock)", 
     ylab="Avg Number of Steps", main="Avg. Number of Steps per Interval\n Weekdays", 
     xaxt="n")
axis(side=1, at=seq(0,288, 12), labels=c(0:24))
plot(mnWeekendInts, type="l", xlab="5-Minute Intervals(24hr clock)", 
     ylab="Avg Number of Steps", main="Avg. Number of Steps per Interval\n Weekends", 
     xaxt="n")
axis(side=1, at=seq(0,288, 12), labels=c(0:24))

```
